Compositional Certified Resource Bounds\ifshort{\vspace{-.15cm}}{}

### 2014-11-08 paper/main.tex
updated the intro

Available techniques for deriving worst-case resource bounds
split in two categories.  Techniques
in the first category derive impressive bounds for
imperative programs, but they are not fully compositional.
This is problematic if one needs to derive global whole-program bounds.
In the second category, techniques derive tight whole-program
bounds on functional programs.
They are highly compositional, scale for large
programs, and work directly on the syntax.  However, despite all
these theoretical and practical advantages, it has been
a long-time open problem to make them work in the
presence of negative integers, mutation, and non-linear
control flow. As a result, they are not suitable for
typical imperative code.
%
\jan{I changed the order here a bit. Good style: First sentence is a
  summary of the following paragraph.}

Notable tools in the first category include
SPEED~\cite{GulwaniMC09}, KoAT~\cite{BrockschmidtEFFG14},
PUBS~\cite{AlbertAGPZ12}, Rank~\cite{AliasDFG10},
and LOOPUS~\cite{SinnZV14}.
They lack compositionality in at least three ways.
First, they all base their analysis on some form of
\emph{ranking-function}.
Since this
is often a local analysis, the loop bounds are
arithmetic expressions that depend on the value of
variables  just before the loop.
This makes it hard to give a
resource bound on sequenced loops or function calls in
terms of the input parameters of a procedure.
Second, while all popular imperative programming languages
provide a function or procedure abstraction, available
tools are not able to abstract their resource behavior
and have to inline the procedure body to perform their
analysis.  Third, when a tool fails to find a resource
bound for a program, the program has to be rewritten.
There is no possibility for sound user interaction with the
tool. For example, there is no concept of manual proofs for
resource bounds and no framework that could support composition
of manually derived bounds with automatically inferred bounds.
One reason for this lack of interaction is that available tools
work on an abstract version of the source program. This
abstracted program has more behaviors than the
original one and is usually expressed in a language that
does not feature memory operations.
%
\jan{@Quentin: What do you mean by ``more behaviors''?}

Tools in the second category are based on type systems for functional
programs and on the potential method of amortized
analysis~\cite{Jost03,HoffmannAH12}.  The first challenge of
imperative programs is already the absence of a fine-grained type system to hook
the analysis on.  Another issue is that the techniques associate
\emph{potential} directly to program variables.  In
imperative programs, this would fail for loops as simple as
\lstinline{for(i=x;i<y;i++)} where $y - i$ decreases, but not $|i|$.
Mutation is yet another challenge.  While functional programs relate
multiple fixed inputs (the typing context) with one output (the
expression typed), imperative programs can change the value of
variables in scope, so the potential of these variables has to be
updated after mutations or, unsound bounds can be derived.
%
\jan{@Quentin: There is a mismatch between the previous two paragraphs.
  In the first one we talk about things that category one techniques
  can't do. In the second one we talk about ``challenges'' for the
  category two. We instead should just say what the state-of-the-art
  can't do. We also have to mention Hofmann/Jost ESOP'06 and Atkey ESOP'10.}

This paper presents a new framework for automatically
deriving resource bounds on C programs.  This new
approach is an attempt to unify the two previous
categories: it solves the compositionality issues of the
imperative world by adapting amortized-analysis based
techniques from the functional world.

Our automated analysis is able to infer linear
resource bounds on C programs with mutually-recursive functions and integer loops.
To our knowledge this is the first technique based on
amortized analysis that is able to derive bounds that depend on negative numbers
and differences of variables.  It is also the first
resource analysis technique for C that deals naturally with
recursive functions, resources that can become available during execution (e.g., when freeing
memory), and sequenced loops.  Compared to more classical
apporaches based on ranking functions, our tool inherits
the benefits of amortized reasoning.  Using only one
simple mechanism, it handles:

\begin{itemize}
\item interactions between \emph{sequential loops or
  function calls} through size changes of variables,
\itemskipIn
\item \emph{nested loops} that influence each other
  with the same set of modified variables,
\itemskipIn
\item and \emph{amortized bounds} as found, for example, in
  the Knuth-Morris-Pratt algorithm for string search.
\end{itemize}
\jan{We can say that our technique is the only one that reduces the problem to
  LP solving. I used that as a selling point in the DARPA-STAC proposal.}

The main innovations that make amortized analysis work
on imperative languages is to base the analysis on a
Hoare-like logic and track multivariate quantities instead
of program variables directly.  This leads to precise bounds
expressed as functions of sizes $|[x, y]| = \max(0, y-x)$ of
intervals.

@@

Give a description of the sections here.                    XXX

\jan{We also mention before that existing imperative tools don't allow
  for user interaction. We should pick this up again and say that we
  can do it. But then we have to mention the PLDI paper. My suggestion:
  \begin{itemize}
  \item mention PLDI'14 in the paragraph about the second category and
    say it has great flexibility for manual bound derivation but
    almost no support for automation
  \item just add another bullet to say that we have the same
    flexibility for manual bound derivation as in PLDI'14 and that
    manual and automatic bounds can be composed
  \end{itemize}
}

>>>>

Available techniques for deriving worst-case resource bounds
fall into two categories.  Techniques
in the first category derive impressive bounds for
numerical imperative programs, but they are not fully compositional.
This is problematic if one needs to derive global whole-program bounds.
In the second category, techniques derive tight whole-program
bounds on functional programs that operate on inductive data structures.
They are highly compositional, scale for large
programs, and work directly on the syntax.  However, despite all
these theoretical and practical advantages, it has been
a long-time open problem to make them work in the
presence of signed integers, mutation, and non-linear
control flow. As a result, they are not suitable for
typical imperative code.

Notable tools in the first category include
SPEED~\cite{GulwaniMC09}, KoAT~\cite{BrockschmidtEFFG14},
PUBS~\cite{AlbertAGPZ12}, Rank~\cite{AliasDFG10},
and LOOPUS~\cite{SinnZV14}.
They lack compositionality in at least three ways.
First, they all base their analysis on some form of
\emph{ranking function}.
Since this
is usually a local analysis and loop bounds are
arithmetic expressions that depend on the value of
variables  just before the loop.
This makes it hard to give a
resource bound on a sequence of loops and procedure calls in
terms of the input parameters of a procedure.
Second, while all popular imperative programming languages
provide a function or procedure abstraction, available
tools are not able to abstract their resource behavior
and have to inline the procedure body to perform their
analysis.  Third, when a tool fails to find a resource
bound for a program, the program has to be rewritten.
There is no possibility for sound user interaction with the
tool. For example, there is no concept of manual proofs for
resource bounds and no framework that could support composition
of manually derived bounds with automatically inferred bounds.
One reason for this lack of interaction is that available tools
work on an abstract version of the source program. This
abstracted program has sometimes more behaviors than the
original one (introduced by non-determinism) and is usually
expressed in a language that does not feature memory operations.

Tools in the second category originate form the potential method of
amortized analysis and type systems for functional
programs~\cite{Jost03,HoffmannAH12}. It has been shown that class
definitions of object oriented programs~\cite{Jost06} and
data-structure predicates of separation logic~\cite{Atkey10} can play
the role of the type system in imperative programs. However, these
techniques fail for numerical imperative programs because they lack
fine-grained data-structure information to hook the analysis on.
Another issue is that the techniques associate \emph{potential}
directly with single program variables.  In imperative programs, this
fails for loops as simple as \lstinline{for(i=x;i<y;i++)} where
$y - i$ decreases, but not $|i|$.  Mutation is yet another problem.
While functional programs relate multiple fixed inputs (the typing
context) with one output (the expression typed), imperative programs
can change the value of variables in scope, so the potential of these
variables has to be updated after mutations to ensure soundness.

This paper presents a new framework for automatically
deriving resource bounds on C programs.  This new
approach is an attempt to unify the two previous
categories: It solves the compositionality issues of techniques
for numerical imperative code by adapting amortized-analysis based
techniques from the functional world.
%
Our automated analysis is able to infer linear
resource bounds on C programs with mutually-recursive functions and integer loops.
%
The resource behavior of functions can be summarized
in a function specification that can be used a every
call-side without access to the function body.
%
To our knowledge this is the first technique based on
amortized analysis that is able to derive bounds that depend on negative numbers
and differences of variables.  It is also the first
resource analysis technique for C that deals naturally with
recursive functions, resources that can become available during execution (e.g., when freeing
memory), and sequenced loops.  Compared to more classical
apporaches based on ranking functions, our tool inherits
the benefits of amortized reasoning.  Using only one
simple mechanism, it handles:

\begin{itemize}
\item interactions between \emph{sequential loops or
  function calls} through size changes of variables,
\itemskipIn
\item \emph{nested loops} that influence each other
  with the same set of modified variables,
\itemskipIn
\item and \emph{amortized bounds} as found, for example, in
  the Knuth-Morris-Pratt algorithm for string search.
\end{itemize}

The main innovation that makes amortized analysis work
on imperative languages is to base the analysis on a
Hoare-like logic and track multivariate quantities instead
of program variables directly.  This leads to precise bounds
expressed as functions of sizes $|[x, y]| = \max(0, y-x)$ of
intervals. A distinctive feature of our analysis system is that
it reduces bound derivation to a linear optimization problem
that can solved by off-the-shelf LP solvers.  This enables the
efficient inference of global bounds for larger programs.

The use of the potential method of amortized analysis makes user
interaction possible in different ways. For one thing, we can directly
combine the new automatic analysis with manually derived bounds in a
previously-developed quantitative Hoare logic~\cite{veristack14} (see
Section XXX).  For another thing, we describe a new mechanism that
allows the separation of quantitative and qualitative verification
(see Section XXX).  Using this mechanism, the user can guide the
analysis by using auxiliary variables and logical assertions that can
be verified by existing qualitative tools such as Hoare logic or
abstract interpretation.  In this way, we can benefit from existing
automation techniques and provide a middle-ground between fully
automatic and fully manual verification to derive. This enables the
semi-automatic inference of non-linear bounds, such as polynomial,
logarithmic, and exponential bounds.

@@

% Give a description of the sections here.                    XXX
% Jan: We don't need that. Usually, people don't read it.

In summary, we make the following \emph{contributions}.

\begin{itemize}
\item We develop the first automatic amortized analysis for C
  programs. It is naturally compositional, tracks size changes of
  variables to derive global bounds, can handle mutually-recursive
  functions, generates resource specifications for functions, derives
  easily verifiable certificates, and handles resources that can
  become available during an execution.
\item We show how to reduce the inference of \emph{linear resource
    bounds} to efficient LP solving.
\item We describe a new method of harnessing existing qualitative
  verification techniques to guide the automatic amortized analysis
  to derive \emph{non-linear} resource bounds.
\item We prove the soundness of the analysis with respect to a
  parametric cost semantics for C programs. The cost model can be
  further customized with function calls (\code{tick(n)}) that
  indicate resource usage. This mechanism is also covered by the
  soundness proof.
\item We present experiments with a prototype implementation on more
  than XXX lines of C code. A detailed comparison shows that our
  prototype is the only tool that can derive global bounds for larger
  C programs while being as powerful as existing tools when deriving
  linear local bounds for tricky loop and recursion patterns.
\end{itemize}
